A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-11 example.tex
Fix first example

\begin{example}\label{ex:string-constraints} Consider the following set of
    string constraints over string variables~$s_1, s_2$ and integer
    variables~$n, i$:
\begin{constraints}
    \item\label{const:s1-in-c-dd} $s_1 \in \Language(\Regex{c\RegexOr{}dd})$.
    \item\label{const:s2-in-b} $s_2 \in \Language(\mathcal{B})$, the language accepted by
    automaton~$\mathcal{B}$ of \cref{fig:aut_b} (ignoring
    the registers~$l_\mathcal{B}$).
    \item\label{const:s1-substring} $s_1 = \Substring(s_2, i, n)$, that is $s_1$ is an
    $n$-length substring of $s_2$ starting at offset $i$.
    \item\label{const:more-inside-than-before} $n > i$, that is what comes
    before $s_1$ in $s_2$ is shorter than $s_1$.
    \item\label{const:something-before-and-after} $i > 0 \land \Length{s_2} -i -n > 0$, that
    is there is at least one character in $s_2$ before and after $s_1$.
\end{constraints}
\end{example}

@@

Intuitively, \cref{ex:string-constraints} is unsatisfiable. By inspecting the
automata, we realise that the path using $\Char{dd}$ in $A'$ is unusable since
are no $\Char{d}$-labelled transitions in~$\mathcal{B}$. This means that $s_1 = \Char{c}$,
and thus $n = 1$. That means that \cref{const:more-inside-than-before} implies
that $i = 0$. However, no path through $B$ passes by a $\Char{c}$ without a
preceding character. Therefore, already
\cref{const:s1-in-c-dd,const:s1-in-c-dd,const:s1-substring,const:more-inside-than-before}
together are unsatisfiable.

@@

We proceed by adding linear constraints requiring transition variables to
represent a flow through their automaton by adding linear constraints stating
that the number of incoming transitions is equal to the number of outgoing ones.
E.g., state $B$ of automaton~$\mathcal{B}$ would have the sum
%
\begin{equation}\label{eq:flow-eqs}
  \cancel{x_{\Tuple{B, \Char{c}, B}}} +
  x_{\Tuple{I_B, \Char{b}, B}} = \cancel{x_{\Tuple{B, \Char{c}, B}}}  +
  x_{\Tuple{B, \Char{c}, F_B}} + x_{\Tuple{B, \Char{b}, I_B}}
\end{equation}
%
where we let
$x_{\Tuple{q, l, c'}}$ refer to the integer variable associated with a
transition from state $q$ to state $q'$ with label $l$. The initial state
of the automaton receives an additional inflow of $1$, and accepting
states have an additional common outflow of $1$. Note that self-loops
cancels themselves out. This means that for automata where a loop can become
unreachable, additional constraints would be required to ensure reachability.
However, being lazy, we put off that work until later.

@@

We then use linear arithmetic reasoning on the register
equations~\eqref{eq:counter-sums} and flow
equations~\eqref{eq:flow-eqs} to simplify the counters associated
with transitions.
As an example, we start with $\mathcal{A}'$. Substituting back
solutions to the equations for $\mathcal{A}'$ to the automaton, we obtain the
automaton in \cref{fig:a_2}, where the notation~$a / t$ now denotes
a transition that accepts letter~$a$ and is taken~$t$ times.
In this case, the transitions of the automaton
can be specified purely in terms of the free
variables we care about representing:  the length of the accepted word ($l_\mathcal{A}$), the
start of the substring ($i$), and the length of the substring ($n$).

@@

Having obtained this representation, we conclude that $1 < n \leq 2$
from
\cref{const:more-inside-than-before,const:something-before-and-after}. The
lower bound directly follows from
\cref{const:more-inside-than-before,const:something-before-and-after},
whereas the upper bound is obtained by the following reasoning. Since all
transition variables are non-negative integers (a transition cannot be
used a negative amount of times), $n \leq 2$ from the $I_A$ to $F_A$
transition. Therefore, it follows that $n=2$, which implies that the
$I_A$ to $F_A$ transition can never be used under these constraints
and that we must use the two $\Char{d}$-labelled transitions, both now
taken $n-1 =1$ times.
% This immediately implies that the problem is
%unsatisfiable, since we know this transition has no correspondance in
%automaton $B$.
%
Substituting the derived value of $n$ and applying similar
reasoning to automaton~$B$, we arrive at the simplified
automata in \cref{fig:propagated}.

@@

In the case~$x_{12} = 0$ we can conclude
that the state~$B$ is now unreachable, which means that also
transitions outgoing from state $B$ are now unusable. Propagation can
therefore infer the additional equations $x_9 = 0$ and $x_{11} = 1$.

\subsubsection{Computing products}\label{sec:intuition:materialise}
After discounting all transitions that can no longer be taken, we are
left with two (small) flat automata, and can compute their product with
relative ease. By putting off computing the product $\mathcal{B} \times \mathcal{A}'$ until after
performing linear reasoning, and using that to prune transitions, we
have computed a smaller product than we would have with an eager
approach. Computing the product, we will immediately notice that the
$\Char{d}$-labelled transition of automaton $\mathcal{A}'$ has no correspondence
in $\mathcal{B}$, leading to an empty product. We can close the proof goal and
backtrack, and will eventually derive that the imposed constraints are
unsatisfiable.

>>>>

\begin{example}\label{ex:string-constraints} Consider the following set of
    string constraints over string variables~$s_1, s_2$ and integer
    variables~$n, i$:
\begin{constraints}
    \item\label{const:s1-in-c-dd} $s_1 \in \Language(\Regex{c\RegexOr{}dd})$.
    \item\label{const:s2-in-b} $s_2 \in \Language(\mathcal{B})$, the language accepted by
    automaton~$\mathcal{B}$ of \cref{fig:aut_b} (ignoring
    the registers~$l_B$).
    \item\label{const:s1-substring} $s_1 = \Substring(s_2, i, n)$, that is $s_1$ is an
    $n$-length substring of $s_2$ starting at offset $i$.
    \item\label{const:more-inside-than-before} $n > i$, that is what comes
    before $s_1$ in $s_2$ is shorter than $s_1$.
    \item\label{const:something-before-and-after} $i > 0 \land \Length{s_2} -i -n > 0$, that
    is there is at least one character in $s_2$ before and after $s_1$.
\end{constraints}
\end{example}

@@

Intuitively, \cref{ex:string-constraints} is unsatisfiable. By inspecting the
automata, we realise that the path using $\Char{dd}$ in $\mathcal{A}'$ is unusable since
are no $\Char{d}$-labelled transitions in~$\mathcal{B}$. This means that $s_1 = \Char{c}$,
and thus $n = 1$. That means that \cref{const:more-inside-than-before} implies
that $i = 0$. However, no path through $B$ passes by a $\Char{c}$ without a
preceding character. Therefore, already
\cref{const:s1-in-c-dd,const:s1-in-c-dd,const:s1-substring,const:more-inside-than-before}
together are unsatisfiable.

@@

We proceed by adding linear constraints requiring transition variables to
represent a flow through their automaton by adding linear constraints stating
that the number of incoming transitions is equal to the number of outgoing ones.
E.g., state $B$ of automaton~$\mathcal{B}$ would have the sum
%
\begin{equation}\label{eq:flow-eqs}
  \cancel{x_{\Tuple{B, \Char{c}, B}}} +
  x_{\Tuple{I_B, \Char{b}, B}} = \cancel{x_{\Tuple{B, \Char{c}, B}}}  +
  x_{\Tuple{B, \Char{c}, F_B}} + x_{\Tuple{B, \Char{b}, I_B}}
\end{equation}
%
where we let
$x_{\Tuple{q, l, q'}}$ refer to the integer variable associated with a
transition from state $q$ to state $q'$ with label $l$. The initial state
of the automaton receives an additional inflow of $1$, and accepting
states have an additional common outflow of $1$. Note that self-loops
cancels themselves out. Therefore, when a loop can become
unreachable, additional constraints are required to ensure consistency.

@@

We then use linear arithmetic reasoning on the register
equations~\eqref{eq:counter-sums} and flow
equations~\eqref{eq:flow-eqs} to simplify the counters associated
with transitions.
As an example, we start with $\mathcal{A}'$. Substituting back
solutions to the equations for $\mathcal{A}'$ to the automaton, we obtain the
automaton in \cref{fig:a_2}, where the notation~$a / t$ now denotes
a transition that accepts letter~$a$ and is taken~$t$ times.
In this case, the transitions of the automaton
can be specified purely in terms of the free
variables we care about representing:  the length of the accepted word ($l_A$), the
start of the substring ($i$), and the length of the substring ($n$).

@@

Having obtained this representation, we conclude that $1 < n \leq 2$
from
\cref{const:more-inside-than-before,const:something-before-and-after}. The
lower bound directly follows from
\cref{const:more-inside-than-before,const:something-before-and-after},
whereas the upper bound is obtained by the following reasoning. Since all
transition variables are non-negative integers (a transition cannot be
used a negative amount of times), $n \leq 2$ from the $I_A$ to $F_A$
transition. Therefore, it follows that $n=2$, which implies that the
$I_A$ to $F_A$ transition can never be used under these constraints
and that we must use the two $\Char{d}$-labelled transitions, both now
taken $n-1 =1$ times.
% This immediately implies that the problem is
%unsatisfiable, since we know this transition has no correspondance in
%automaton $B$.
%
Substituting the derived value of $n$ and applying similar
reasoning to automaton~$\mathcal{B}$, we arrive at the simplified
automata in \cref{fig:propagated}.

@@

In the case~$x_{12} = 0$, we can conclude
that the state~$B$ is now unreachable, which means that its outgoing transitions are now unusable. Propagation can
therefore infer the additional equations $x_9 = 0$ and $x_{11} = 1$.

\subsubsection{Computing products}\label{sec:intuition:materialise}
After discounting all transitions that can no longer be taken, we are
left with two (small) flat automata, and can compute their product with
relative ease. By putting off computing the product $\mathcal{B} \times \mathcal{A}'$ until after
performing linear reasoning, and using that to prune transitions, we
have computed a smaller product than we would have with an eager
approach. Computing the product, we will immediately notice that the
$\Char{d}$-labelled transition of automaton $\mathcal{A}'$ has no correspondence
in $\mathcal{B}$, leading to an empty product. We can close the proof goal and
backtrack, and we will eventually derive that the imposed constraints are
unsatisfiable by repeating the same process on the other branch.
